---
title: "Symptom Prediction in Schizophrenia"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Predict Depression Levels

### Load and format data
```{r}
szdat <- read.csv("feature_matrix_anomalies_trunc3.csv")

# create vector of outcome variable (depression)
Y <- szdat$Depression
# create matrix of predictors (15 mobility measures)
X <- szdat[,9:23]

# remove observations with NA's
rmX <- which(apply(X,1,function(yy) length(which(is.na(yy)))>0))
rmY <- which(is.na(Y))
rmXY <- unique(c(rmX,rmY))
lab <- szdat[-rmXY,1]
Y <- Y[-rmXY]
X <- scale(X[-rmXY,])   # scale predictors
```

### Split data into training and testing  
The last observation per subject is withheld from training and used for calculating mean squared prediction error (MSPE).
```{r}
# m: number of subjects
m <- length(unique(lab))

# n: number of observations per subject
n <- sapply(1:m, function(i) length(which(lab==unique(lab)[i])))

# get index of last observation for each subject
last.obs <- sapply(1:length(n), function(i) sum(n[1:i]))

# training data
Y.train <- Y[-last.obs]
X.train <- X[-last.obs,]
lab.train <- lab[-last.obs]

# testing data (last observation per subject)
Y.test <- Y[last.obs]
X.test <- X[last.obs,]
lab.test <- lab[last.obs]

# set up predictor matrix for maity-pal method
subject.indicator <- matrix(rep(0, m*sum(n)), nrow = sum(n), ncol = m)
for (i in 1:m) {
  subject.indicator[(sum(n[0:(i-1)])+1):(sum(n[1:i])),i] <- rep(1, n[i])
}
X.mp <- cbind(X, subject.indicator)
X.mp.train <- X.mp[-last.obs,]
X.mp.test <- X.mp[last.obs,]
```

### Run four models  
MSPE is averaged across 5 models fitted with different starting weights and biases for GNMM, ANN, and MP. Due to the larger network size in the MP model, training is done with a large penalty term that decreases every 10 epochs.
```{r}
# source functions for models
source('Network_Functions.R')

# network settings
nepochs <- 50
hidnodes1 <- 3
hidnodes2 <- 2

# run 1-layer GNMM
set.seed(12044)
mspe.gnmm1 <- rep(NA,5)
for (i in 1:5) {
  m1 <- gnmm.sgd(formula = Y.train ~ X.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                  nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.005, act_fun = 'relu', nepochs = nepochs, incl_ranef = TRUE)
  pred1 <- gnmm.predict(new_data = X.test, id = lab.test, gnmm.fit = m1)
  mspe.gnmm1[i] <- mean((Y.test-pred1)^2)
}

# run 2-layer GNMM
set.seed(12045)
mspe.gnmm2 <- rep(NA,5)
for (i in 1:5) {
  m2 <- gnmm.sgd(formula = Y.train ~ X.train + (1|lab.train), family = 'gaussian', penalization = 0.002,
                  nodes1 = hidnodes1, nodes2 = hidnodes2, step_size = 0.005, act_fun = 'relu', nepochs = nepochs, incl_ranef = TRUE)
  pred2 <- gnmm.predict(new_data = X.test, id = lab.test, gnmm.fit = m2)
  mspe.gnmm2[i] <- mean((Y.test-pred2)^2)
}

# run ANN
set.seed(12045)
mspe.ann <- rep(NA,5)
for (i in 1:5) {
  m3 <- gnmm.sgd(formula = Y.train ~ X.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                  nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.005, act_fun = 'relu', nepochs = nepochs, incl_ranef = FALSE)
  pred3 <- gnmm.predict(new_data = X.test, id = lab.test, gnmm.fit = m3)
  mspe.ann[i] <- mean((Y.test-pred3)^2)
}

# run GLMM with main effects
p <- ncol(X.train)
glmm.main <- glmer(Y.train ~ X.train + (1|lab.train), family = 'gaussian')
coef.main <- matrix(summary(glmm.main)$coefficients[2:(p+1),1])
fef.main <- X.test%*%coef.main + rep(matrix(summary(glmm.main)$coefficients[1,1]),m)     # fixed effect
ref.main <- unlist(unname(as.vector(c(ranef(glmm.main)$lab.train))))                     # random effect
pred4 <- fef.main + ref.main

# run GLMM with main effects + 3 interaction terms
glmm.3int <- glmer(Y.train ~ X.train + X.train[,4]*X.train[,15] + X.train[,3]*X.train[,15] + X.train[,5]*X.train[,13] + 
                     (1|lab.train), family = 'gaussian')
coef.3int <- matrix(summary(glmm.3int)$coefficients[2:(p+3+1),1])
Xtest.3int <- cbind(X.test, X.test[,4]*X.test[,15], X.test[,3]*X.test[,15], X.test[,5]*X.test[,13])        # matrix with interactions
fef.3int <- Xtest.3int%*%coef.3int + rep(matrix(summary(glmm.3int)$coefficients[1,1]),m)      # fixed effect
ref.3int <- unlist(unname(as.vector(c(ranef(glmm.3int)$lab.train))))                          # random effect
pred5 <- fef.3int + ref.3int

# run GLMM with main effects + 5 interaction terms
glmm.5int <- glmer(Y.train ~ X.train + X.train[,4]*X.train[,15] + X.train[,3]*X.train[,15] + X.train[,5]*X.train[,13] + 
                     X.train[,7]*X.train[,13] + X.train[,5]*X.train[,12] + (1|lab.train), family = 'gaussian')
coef.5int <- matrix(summary(glmm.5int)$coefficients[2:(p+5+1),1])
Xtest.5int <- cbind(X.test, X.test[,4]*X.test[,15], X.test[,3]*X.test[,15], X.test[,5]*X.test[,13], 
                    X.test[,7]*X.test[,13], X.test[,5]*X.test[,12])           # matrix with interactions
fef.5int <- Xtest.5int%*%coef.5int + rep(matrix(summary(glmm.5int)$coefficients[1,1]),m)      # fixed effect
ref.5int <- unlist(unname(as.vector(c(ranef(glmm.5int)$lab.train))))                          # random effect
pred6 <- fef.5int + ref.5int

# run MP
# 4th iteration does not converge so model is run 6 times and 4th run is dropped
set.seed(12047)
mspe.mp <- rep(NA,6)
for (i in 1:6) {
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 1,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.75025,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.50050,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.25075,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  pred7 <- gnmm.predict(new_data = X.mp.test, id = lab.test, gnmm.fit = m5)
  mspe.mp[i] <- mean((Y.test-pred7)^2) 
}
```

### Compare MSPE from 4 models  
```{r}
# 1-layer GNMM
mean(mspe.gnmm1)              # 0.428   

# 2-layer GNMM
mean(mspe.gnmm2)              # 0.422  

# ANN
mean(mspe.ann)                # 1.076
 
# GLMM (main effects)
mean((Y.test-pred4)^2)        # 0.546

# GLMM (3 interactions)
mean((Y.test-pred5)^2)        # 0.609

# GLMM (5 interactions)
mean((Y.test-pred6)^2)        # 0.588

# MP
mean(mspe.mp, na.rm = TRUE)   # 1.218
```


## Predict Anxiety Levels

### Load and format data
```{r}
szdat <- read.csv("feature_matrix_anomalies_trunc3.csv")

# create vector of outcome variable (anxiety)
Y <- szdat$Anxiety
# create matrix of predictors (15 mobility measures)
X <- szdat[,9:23]

# remove observations with NA's
rmX <- which(apply(X,1,function(yy) length(which(is.na(yy)))>0))
rmY <- which(is.na(Y))
rmXY <- unique(c(rmX,rmY))
lab <- szdat[-rmXY,1]
Y <- Y[-rmXY]
X <- scale(X[-rmXY,])   # scale predictors
```

### Split data into training and testing  
The last observation per subject is withheld from training and used for calculating mean squared prediction error (MSPE).
```{r}
# m: number of subjects
m <- length(unique(lab))

# n: number of observations per subject
n <- sapply(1:m, function(i) length(which(lab==unique(lab)[i])))

# get index of last observation for each subject
last.obs <- sapply(1:length(n), function(i) sum(n[1:i]))

# training data
Y.train <- Y[-last.obs]
X.train <- X[-last.obs,]
lab.train <- lab[-last.obs]

# testing data (last observation per subject)
Y.test <- Y[last.obs]
X.test <- X[last.obs,]
lab.test <- lab[last.obs]

# set up predictor matrix for maity-pal method
subject.indicator <- matrix(rep(0, m*sum(n)), nrow = sum(n), ncol = m)
for (i in 1:m) {
  subject.indicator[(sum(n[0:(i-1)])+1):(sum(n[1:i])),i] <- rep(1, n[i])
}
X.mp <- cbind(X, subject.indicator)
X.mp.train <- X.mp[-last.obs,]
X.mp.test <- X.mp[last.obs,]
```

### Run four models  
MSPE is averaged across 5 models fitted with different starting weights and biases for GNMM, ANN, and MP. Due to the larger network size in the MP model, training is done with a large penalty term that decreases every 10 epochs.
```{r}
# source functions for models
source('Network_Functions.R')

# network settings
nepochs <- 50
hidnodes1 <- 3
hidnodes2 <- 2

# Run 1-layer GNMM
# 1st and 6th iterations do not converge so model is run 7 times and 2 runs are dropped
set.seed(12046)
mspe.gnmm1 <- rep(NA,7)
for (i in 1:7) {
  m1 <- gnmm.sgd(formula = Y.train ~ X.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                  nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.01, act_fun = 'relu',nepochs = nepochs, incl_ranef = TRUE)
  pred1 <- gnmm.predict(new_data = X.test, id = lab.test, gnmm.fit = m1)
  mspe.gnmm1[i] <- mean((Y.test-pred1)^2)
}

# Run 2-layer GNMM
set.seed(12046)
mspe.gnmm2 <- rep(NA,5)
for (i in 1:5) {
  m2 <- gnmm.sgd(formula = Y.train ~ X.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                  nodes1 = hidnodes1, nodes2 = hidnodes2, step_size = 0.01, act_fun = 'relu',nepochs = nepochs, incl_ranef = TRUE)
  pred2 <- gnmm.predict(new_data = X.test, id = lab.test, gnmm.fit = m2)
  mspe.gnmm2[i] <- mean((Y.test-pred2)^2)
}

# Run ANN
# 4th iteration does not converge so model is run 6 times and 4th run is dropped
set.seed(12045)
mspe.ann <- rep(NA,6)
for (i in 1:6) {
  m3 <- gnmm.sgd(formula = Y.train ~ X.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                  nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.01, act_fun = 'relu', nepochs = nepochs, incl_ranef = FALSE)
  pred3 <- gnmm.predict(new_data = X.test, id = lab.test, gnmm.fit = m3)
  mspe.ann[i] <- mean((Y.test-pred3)^2)
}

# Run GLMM with main effects
p <- ncol(X.train)
glmm.main <- glmer(Y.train ~ X.train + (1|lab.train), family = 'gaussian')
coef.main <- matrix(summary(glmm.main)$coefficients[2:(p+1),1])
fef.main <- X.test%*%coef.main + rep(matrix(summary(glmm.main)$coefficients[1,1]),m)     # fixed effect
ref.main <- unlist(unname(as.vector(c(ranef(glmm.main)$lab.train))))                     # random effect
pred4 <- fef.main + ref.main

# Run GLMM with main effects + 3 interaction terms
glmm.3int <- glmer(Y.train ~ X.train + X.train[,2]*X.train[,5] + X.train[,5]*X.train[,13] + X.train[,5]*X.train[,11] + 
                     (1|lab.train), family = 'gaussian')
coef.3int <- matrix(summary(glmm.3int)$coefficients[2:(p+3+1),1])
Xtest.3int <- cbind(X.test, X.test[,2]*X.test[,5], X.test[,5]*X.test[,13], X.test[,5]*X.test[,11])        # matrix with interactions
fef.3int <- Xtest.3int%*%coef.3int + rep(matrix(summary(glmm.3int)$coefficients[1,1]),m)         # fixed effect
ref.3int <- unlist(unname(as.vector(c(ranef(glmm.3int)$lab.train))))                             # random effect
pred5 <- fef.3int + ref.3int

# Run GLMM with main effects + 5 interaction terms
glmm.5int <- glmer(Y.train ~ X.train + X.train[,2]*X.train[,5] + X.train[,5]*X.train[,13] + X.train[,5]*X.train[,11] + 
                     X.train[,2]*X.train[,4] + X.train[,3]*X.train[,12] + (1|lab.train), family = 'gaussian')
coef.5int <- matrix(summary(glmm.5int)$coefficients[2:(p+5+1),1])
Xtest.5int <- cbind(X.test, X.test[,2]*X.test[,5], X.test[,5]*X.test[,13], X.test[,5]*X.test[,11], 
                    X.test[,2]*X.test[,4], X.test[,3]*X.test[,12])            # matrix with interactions
fef.5int <- Xtest.5int%*%coef.5int + rep(matrix(summary(glmm.5int)$coefficients[1,1]),m)     # fixed effect
ref.5int <- unlist(unname(as.vector(c(ranef(glmm.5int)$lab.train))))                         # random effect
pred6 <- fef.5int + ref.5int

# Run MP
# 4th iteration does not converge so model is run 6 times and 4th run is dropped
set.seed(12047)
mspe.mp <- rep(NA,6)
for(i in 1:6) {
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 1,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.75025,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = .50050,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.25075,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  m5 <- gnmm.sgd(formula = Y.train ~ X.mp.train + (1|lab.train), family = 'gaussian', penalization = 0.001,
                 nodes1 = hidnodes1, nodes2 = NULL, step_size = 0.05, act_fun = 'relu', nepochs = 10, incl_ranef = FALSE,
                 weights = m5$weights, biases = m5$biases)
  pred7 <- gnmm.predict(new_data = X.mp.test, id = lab.test, gnmm.fit = m5)
  mspe.mp[i] <- mean((Y.test-pred7)^2)
}
```

### Compare MSPE from 4 models  
```{r}
# 1-layer GNMM
mean(mspe.gnmm1, na.rm = TRUE)    # 0.062   

# 2-layer GNMM
mean(mspe.gnmm2)    # 0.061   

# ANN
mean(mspe.ann, na.rm = TRUE)      # 0.413

# GLMM (main effects)
mean((Y.test-pred4)^2)    # 0.068

# GLMM (3 interactions)
mean((Y.test-pred5)^2)    # 0.098

# GLMM (5 interactions)
mean((Y.test-pred6)^2)    # 0.116

# MP
mean(mspe.mp, na.rm=TRUE)    # 0.518
```


### Significant Location Entropy Plot
```{r}
# load data
szdat <- read.csv("feature_matrix_anomalies_trunc3.csv")

# create vector of outcome variable (anxiety)
Y <- szdat$Anxiety
# create matrix of predictors (15 mobility measures)
X <- szdat[,9:23]

# remove observations with NA's
rmX <- which(apply(X,1,function(yy) length(which(is.na(yy)))>0))
rmY <- which(is.na(Y))
rmXY <- unique(c(rmX,rmY))
lab <- szdat[-rmXY,1]
Y <- Y[-rmXY]
X <- X[-rmXY,]

# m: number of subjects
m <- length(unique(lab))

# n: number of observations per subject
n <- sapply(1:m, function(i) length(which(lab==unique(lab)[i])))

# get index of last observation for each subject
last.obs <- sapply(1:length(n), function(i) sum(n[1:i]))

# set up ordinal anxiety vector
med.anx <- c(median(Y[1:last.obs[1]]), sapply(2:m, function(i) median(Y[(last.obs[i-1]+1):(last.obs[i])])))
med.anx.expanded <- rep(med.anx, times=n)
ord.anxiety <- rep(NA, length(Y))
for (j in 1:length(Y)) {
  if(Y[j]==med.anx.expanded[j]){
    ord.anxiety[j] <- 0
  } else if (Y[j]>med.anx.expanded[j]){
    ord.anxiety[j] <- 1
  } else if (Y[j]<med.anx.expanded[j]){
    ord.anxiety[j] <- -1
  }
}

# plot
p1 <- ggplot(NULL, aes(x = X[,12], y = ord.anxiety)) + geom_point(alpha = 0.15) + xlim(0,0.75) + 
  geom_smooth(method = 'loess', color = 'red', alpha = 0.4, size= 1.5) + ylab('Scaled Anxiety') +
  xlab('Significant Location Entropy') + theme_classic() + theme(panel.border = element_rect(colour = "black", fill=NA))
p1
```
